---
title: "Chronological data from XRONOS in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chronological data from XRONOS in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[XRONOS](https://xronos.ch) is a worldwide database of chronological data (such as radiocarbon or dendrochronological dates) from archaeological contexts.
The xronos package allows you to retrieve data from XRONOS directly using R.
It also provides tools for inspecting and transforming this data for analysis with commonly-used packages.

This vignette introduces the main features of the xronos package.
It covers:

* Querying and retrieving chronological data from XRONOS with `chron_data()`
* Calibrating radiocarbon dates with xronos and c14
* Mapping chronological data with xronos and sf

```{r setup}
library(xronos)
```


## Get chronological data

`chron_data()` provides an R interface to [XRONOS' REST API](https://xronos.ch/api) for retrieving chronological data.
Without further parameters, it will request all available data from XRONOS.

```{r get-all-data, eval=FALSE}
chron_data()
```

Since this takes some time, you will be prompted to confirm this request.
Use `chron_data(.everything = TRUE)` to confirm that you do want to download all records from XRONOS and suppress this prompt.

In practice, you'll probably want to narrow down the dataset with a search query.
You can use the following parameters, passed as arguments to `chron_data()`, to filter the results:

* `labnr`
* `site`
* `site_type`
* `country`
* `feature`
* `material`
* `species`

You can combine any number of parameters, and each parameter can take either a single value or a vector of values that should be included in the results.

For example, to get all radiocarbon dates from Switzerland on either charcoal or bone:

```{r}
swiss_data <- chron_data(country = "Switzerland", material = c("charcoal", "bone"))
```

The results of the query are returned as a table of records (a `tibble` if tibble is installed, otherwise a data frame):

```{r}
swiss_data
```


## Transformation into a spatial dataset and mapping

xronos has an inbuilt function `chron_as_sf()` for turning to received data into a special dataset for the package `sf`. Keep in mind, that this only works, if `sf` is installed on your machine.

```{r}
swiss_data.sf <- chron_as_sf(swiss_data)
```

The resulting special dataset can be used easily to map the results of the query, for example with the inbuild functions of `sf`:

```{r}
plot(sf::st_geometry(swiss_data.sf))
```

If you have map for example `ggmap` installed, you can get much more elaborated mapping directly:

```{r}
library(ggmap)
bounding <- sf::st_bbox(swiss_data.sf)
names(bounding) <- c("left","bottom","right","top") 

map <- get_map(location = bounding, zoom = 8, maptype = "terrain", source = "stamen")

# Just plotting, the geom_sf layer is not aligned properly:
ggmap(map) + 
  geom_sf(data = swiss_data.sf, inherit.aes = FALSE, color = "red")
```

## Calibration

To calibrate to receive data, you can directly use the columns `bp` and `std` and use the calibration environment of your choice. Additionally, it might be helpful to use the `labnr` column to provide an id for your data. We suggest that you might use the `c14` package of Joe Roe. Other options might be for example `rcarbon`. :

```{r}
library(rcarbon)

swiss_data.calibrated <- calibrate(x=swiss_data$bp,
                                   errors=swiss_data$std,
                                   ids = swiss_data$labnr,
                                   verbose = F)
```

After that, you can visualise the calibration result with the appropriate tools from your selected package:

```{r}
multiplot(swiss_data.calibrated[1:5])
```

