---
title: "Chronological data from XRONOS in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chronological data from XRONOS in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[XRONOS](https://xronos.ch) is a worldwide database of chronological data (such as radiocarbon or dendrochronological dates) from archaeological contexts.
The xronos package allows you to retrieve data from XRONOS directly using R.
It also provides tools for inspecting and transforming this data for analysis with commonly-used packages.

This vignette introduces the main features of the xronos package.
It covers:

* Querying and retrieving chronological data from XRONOS with `chron_data()`
* Calibrating radiocarbon dates with xronos and c14
* Mapping chronological data with xronos and sf

```{r setup}
library(xronos)
```

# Data Retrieval

The primary interface for retrieving data with xronos is the command `chron_data()`, which requests data via XRONOS' REST API.
You can access all data using:

```{r, eval=FALSE}
# We do not run this command, since it takes some time
all_data <- chron_data()
```

But it is more advisable to be selective, and used to query options to specify which data are really requested. Currently you can specify the following search criteria:

* "labnr"
* "site"
* "site_type"
* "country"
* "feature"
* "material"
* "species"

For example, for retrieving all dates from Switzerland from either charcoal or bone, you can use the following command:

```{r}
swiss_data <- chron_data(country = "Switzerland", material = c("charcoal", "bone"))
```

As you can see from the output of the variable, the resulting data are formatted as a tibble.

```{r}
swiss_data
```

# Transformation into a spatial dataset and mapping

xronos has an inbuilt function `chron_as_sf()` for turning to received data into a special dataset for the package `sf`. Keep in mind, that this only works, if `sf` is installed on your machine.

```{r}
swiss_data.sf <- chron_as_sf(swiss_data)
```

The resulting special dataset can be used easily to map the results of the query, for example with the inbuild functions of `sf`:

```{r}
plot(sf::st_geometry(swiss_data.sf))
```

If you have map for example `ggmap` installed, you can get much more elaborated mapping directly:

```{r}
library(ggmap)
bounding <- sf::st_bbox(swiss_data.sf)
names(bounding) <- c("left","bottom","right","top") 

map <- get_map(location = bounding, zoom = 8, maptype = "terrain", source = "stamen")

# Just plotting, the geom_sf layer is not aligned properly:
ggmap(map) + 
  geom_sf(data = swiss_data.sf, inherit.aes = FALSE, color = "red")
```

# Calibration

To calibrate to receive data, you can directly use the columns `bp` and `std` and use the calibration environment of your choice. Additionally, it might be helpful to use the `labnr` column to provide an id for your data. We suggest that you might use the `c14` package of Joe Roe. Other options might be for example `rcarbon`. :

```{r}
library(rcarbon)

swiss_data.calibrated <- calibrate(x=swiss_data$bp,
                                   errors=swiss_data$std,
                                   ids = swiss_data$labnr,
                                   verbose = F)
```

After that, you can visualise the calibration result with the appropriate tools from your selected package:

```{r}
multiplot(swiss_data.calibrated[1:5])
```

