---
title: "Chronological data from XRONOS in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chronological data from XRONOS in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[XRONOS](https://xronos.ch) is a worldwide database of chronological data (such as radiocarbon or dendrochronological dates) from archaeological contexts.
The xronos package allows you to retrieve data from XRONOS directly using R.
It also provides tools for inspecting and transforming this data for analysis with commonly-used packages.

This vignette introduces the main features of the xronos package.
It covers:

* Querying and retrieving chronological data from XRONOS with `chron_data()`
* Calibrating radiocarbon dates with xronos and c14
* Mapping chronological data with xronos and sf

```{r setup}
library(xronos)
```


## Get chronological data

`chron_data()` provides an R interface to [XRONOS' REST API](https://xronos.ch/api) for retrieving chronological data.
Without further parameters, it will request all available data from XRONOS.

```{r get-all-data, eval=FALSE}
chron_data()
```

Since this takes some time, you will be prompted to confirm this request.
Use `chron_data(.everything = TRUE)` to confirm that you do want to download all records from XRONOS and suppress this prompt.

In practice, you'll probably want to narrow down the dataset with a search query.
You can use the following parameters, passed as arguments to `chron_data()`, to filter the results:

* `labnr`
* `site`
* `site_type`
* `country`
* `feature`
* `material`
* `species`

You can combine any number of parameters, and each parameter can take either a single value or a vector of values that should be included in the results.

For example, to get all radiocarbon dates from Switzerland on either charcoal or bone:

```{r get-swiss-data}
chron_ch <- chron_data(country = "Switzerland", material = c("charcoal", "bone"))
```

The results of the query are returned as a table of records (a `tibble` if tibble is installed, otherwise a data frame):

```{r show-swiss-data}
chron_ch
```


## Map chronological data

Wherever possible, XRONOS includes the geographic location of chrons, stored as latitude and longitude coordinates.
The convenience function `chron_as_sf()` converts these coordinates into a 'simple features' geometry column for use with the sf package:

```{r as-sf}
if (!requireNamespace("sf")) install.packages("sf")
library(sf)

chron_as_sf(chron_ch)
```

The result is an `sf` object, which includes the original data, plus a geometry column representing the point coordinates and information on the coordinate reference system used.

Change the `crs` parameter to transform the geometries into another coordinate reference system.
This can be anything understood by `sf::st_crs()`).
For example, to use the Swiss National Grid ([EPSG:2056](https://epsg.io/2056)):

```{r as-sf-projected}
chron_as_sf(chron_ch, crs = 2056)
```

Sometimes you might see a warning about rows with missing or invalid coordinates being dropped.
This is because it is currently not possible to include rows without a geometry in a simple features table.
To suppress the warning—and clarify your intent—you may want to explicitly remove these before calling `chron_as_sf()`:

```{r as-sf-complete}
chron_ch |>
  chron_drop_na_coords() |>
  chron_as_sf(crs = 2056) ->
  chron_ch
```

sf and related packages offer many options for manipulating, analysing, and mapping spatial data.
As a simple example, we can plot the location of our dates using their natural projection (here we use `sf::st_geometry()` to plot only the points themselves, not their data attributes):

```{r}
plot(sf::st_geometry(chron_ch), graticule = TRUE, axes = TRUE)
```

See the [sf documentation](https://r-spatial.github.io/sf/articles/sf5.html) for more options for plotting simple features.

`sf` objects generally act like data frames, but if necessary you can explicitly restore the plain table of data with `sf::st_drop_geometry()`:

```{r undo-sf}
chron_ch <- st_drop_geometry(chron_ch)
```

Note that this does *not* restore any records that were dropped by `chron_as_sf()`. 

## Calibration

To calibrate to receive data, you can directly use the columns `bp` and `std` and use the calibration environment of your choice. Additionally, it might be helpful to use the `labnr` column to provide an id for your data. We suggest that you might use the `c14` package of Joe Roe. Other options might be for example `rcarbon`. :

```{r}
library(rcarbon)

swiss_data <- chron_ch

swiss_data.calibrated <- calibrate(x=swiss_data$bp,
                                   errors=swiss_data$std,
                                   ids = swiss_data$labnr,
                                   verbose = F)
```

After that, you can visualise the calibration result with the appropriate tools from your selected package:

```{r}
multiplot(swiss_data.calibrated[1:5])
```

